---
title: "MOUD in Florida Medicaid Population"
author: 'Michael Q. Maguire, MS'
date: 'February 17, 2022'
output: github_document
---

## Packages Used

```{r packageLoad, message = FALSE}
library(tidyverse)
library(data.table)
library(rvest)
library(hrbrthemes)
```

```{r sdudFile, include = FALSE}
file_location <- 'C:/Users/michaelqmaguire2/Dropbox (UFL)/01_projects/medicaid-drug-utilization/medicaid-drug-utilization/data/clean/03_sdud-redbook-final.csv'
```

## Read in the SDUD data set

```{r sdudLoad}
sdud <- fread(
  file_location,
  colClasses = c('proper_ndc' = 'character')
)
```

## Subset to Florida Only and Create MOUD Flags

Variable `moudFlag` represents whether the brand name has a pattern match with the below values.
Variable `medication` represents the categorization according to AG.

```{r sdudFL}
sdudFL <- sdud[
  i = (state == 'FL' & data.table::between(year, 2016, 2021)),
  j = .(utilization.type, state, year, quarter, suppression, gennme, prodnme, numberrx)
][
  , 
  `:=` (moudFlag = ifelse(grepl(x = prodnme, pattern = "bupreno|suboxone|zubsolv|cassipa|bunavail|subutex", ignore.case = TRUE) == TRUE, 1, 0),
        medication = case_when(
          grepl(x = prodnme, pattern = 'bunavail', ignore.case = TRUE) == TRUE ~ "Bunavail",
          grepl(x = prodnme, pattern = 'subutex', ignore.case = TRUE) == TRUE ~ 'Subutex',
          grepl(x = prodnme, pattern = 'suboxone', ignore.case = TRUE) == TRUE ~ 'Suboxone',
          grepl(x = prodnme, pattern = 'cassipa', ignore.case = TRUE) == TRUE ~ 'Cassipa',
          grepl(x = prodnme, pattern = 'zubsolv', ignore.case = TRUE) == TRUE ~ 'Zubsolv',
          grepl(x = prodnme, pattern = 'buprenorphine', ignore.case = TRUE) == TRUE ~ 'Buprenorphine',
          TRUE ~ 'Not an OUD Drug'
          ),
        yearQuarter = zoo::as.yearqtr(format(paste0(year, quarter)), "%Y%q")
        )
]

setorder(sdudFL, state, yearQuarter, gennme, prodnme, suppression)

sdudFL |> janitor::tabyl(moudFlag)
sdudFL |> janitor::tabyl(medication)
sdudFL |> janitor::tabyl(medication, moudFlag)
```

## Subset to only OUD medications and aggregate.

```{r sdudFlAgg}
sdudFLAggregate <- sdudFL[
  i = moudFlag == 1 & suppression == 'F',
  j = .(totalRX = sum(numberrx)),
  by = .(yearQuarter, medication)
]

setorder(sdudFLAggregate, medication, yearQuarter)

sdudFLAggregate
```

## Create table containing all years, quarters, and drugs.

```{r yearQuarterTable}
yearQuarter <- unique(sdudFL$yearQuarter, by = yearQuarter)

allDrugs <- unique(sdudFLAggregate$medication, by = medication)

cartesianYearQuarterDrug <- CJ(yearQuarter, allDrugs)

cartesianYearQuarterDrug
```

## Full Join with the above table with the `sdudFLAggregate` to get values that don't exist for a given yearQuarter.

```{r getAllQuarters}
yearQuarterAllFLAggregate <- merge.data.table(
  cartesianYearQuarterDrug,
  sdudFLAggregate,
  by.x = c('yearQuarter', 'allDrugs'),
  by.y = c('yearQuarter', 'medication'),
  all.x = TRUE
)

sum(is.na(yearQuarterAllFLAggregate$totalRX))

yearQuarterAllFLAggregate[is.na(totalRX), totalRX := 0]

yearQuarterAllFLAggregate

sum(is.na(yearQuarterAllFLAggregate$totalRX))
```

# Plotting time!

```{r plottingOUD, fig.width = 20, fig.height = 12}
plotRawDrugCount <- yearQuarterAllFLAggregate |>
  ggplot() +
  geom_point(aes(x = as.factor(yearQuarter), y = totalRX, color = allDrugs), size = 2) +
  geom_line(aes(x = as.factor(yearQuarter), y = totalRX, group = allDrugs, color = allDrugs), size = 1) +
  scale_color_manual(values = c("Bunavail" = "grey", "Buprenorphine" = "#FA841A", "Suboxone" = "#004DA9", "Zubsolv" = "chartreuse4")) +
  ggtitle("Florida MOUD Raw Prescription Drug Counts") +
  xlab("Year-Quarter") +
  ylab("Total Prescriptions") +
  theme_ipsum_rc() + 
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14, hjust = 0.5, vjust = -1),
    axis.title.y = element_text(size = 14, hjust = 0.5),
    legend.position = 'bottom',
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.key.size = unit(3, 'line')
  ) +
  labs(color = 'Drug Name')

plotRawDrugCount
```

## Iteratively download files from Medicaid


```{r fileDL, eval = FALSE, message = FALSE}
achaURL <- 'https://ahca.myflorida.com/medicaid/Finance/data_analytics/enrollment_report/' # url itself
achaRead <- read_html(achaURL) # read in the HTML
achaFiles <- achaRead |> html_nodes('a') |> html_attr('href') |> str_subset('ENR') # extract needed elements
achaTrimmed <- achaFiles |> str_extract(pattern = "docs.*") # extract anything after docs
achaYears <- str_subset(achaTrimmed, pattern = "2016|2017|2018|2019|2020|2021") # extract files with desired years
achaNames <- str_extract(achaYears, pattern = "ENR_.*") # get names for naming file on download.file
achaFilesDLURL <- paste0(achaURL, achaYears) # paste url and name together for download.file

# Iteration occurs here

map2(
  achaFilesDLURL,
  achaNames,
  ~download.file(.x, destfile = paste0("./data/raw/", .y), mode = 'wb')
)

# Renaming two pesky files.

file.rename("./data/raw/ENR_February_2016.xls", "./data/raw/ENR_201602.xls")
file.rename("./data/raw/ENR_January_2016.xls", "./data/raw/ENR_201601.xls")
```

## Showing files in directory

```{r fileList}
files <- list.files(path = './data/raw/', full.names = TRUE)
files
```

## Iteratively determine how many columns exist in each file

```{r colCheck, message = FALSE}
numColumnsflMedicaid2016to2021 <- 
  files |>
  map_dfr(
    ~(data.frame
      (number_columns = ncol(
        readxl::read_excel(
          path = .,
          skip = 6,
          sheet = 'AGEGROUP'
        )
      ),
      id = .)
    )
  )

numColumnsflMedicaid2016to2021

numColumnsflMedicaid2016to2021 |>
  distinct(number_columns, .keep_all = TRUE)
```

We can see above that the files do not have the same lengths across the `AGEGROUP` sheet.

I looked at the files that had different lengths and determined how to accommodate the code to each sheet.

I also found that the row value of `TOTAL OF MEDICAID` contained the numbers we needed.

```{r nameVector}
namesAcha18 <- c(
  'plan_name', 'type', 'region', 'blank', 'tanf_children_0_18', 'tanf_children_19_20', 'tanf_adult_21_over',
  'tanf_total', 'blank2', 'medicaid_only_0_18', 'medicaid_only_19_20', 'medicaid_21_over', 'duals_0_18', 'duals_19_20',
  'duals_21_over', 'medicaid_duals_total', 'blank3', 'children_percent'
)

namesAcha20 <- c(
  'plan_name', 'type', 'region', 'blank', 'tanf_children_0_18', 'tanf_children_19_20', 'tanf_adult_21_over',
  'tanf_total', 'blank2', 'medicaid_only_0_18', 'medicaid_only_19_20', 'medicaid_21_over', 'duals_0_18', 'duals_19_20',
  'duals_21_over', 'medicaid_duals_total', 'blank3' , 'tanf_medicaid_duals_total', 'blank4', 'children_percent'
)
```

## Extracting information from ACHA files

Next, I extracted the file paths containing 20 columns, read them in, combined them, and selected the only row we needed.

```{r extract20Cols}
colsGE20 <- numColumnsflMedicaid2016to2021$id[numColumnsflMedicaid2016to2021$number_columns == 20]

# Extract all 20 column files.

acha20 <- setNames(
  lapply(
    colsGE20,
    FUN = function(x) readxl::read_excel(
      path = x,
      sheet = 'AGEGROUP',
      skip = 6,
      col_names = namesAcha20,
      col_types = 'text'
      )
    ),
  colsGE20
  )

# Combine, and keep only necessary row. Extract date information from filename.

flMedicaidGE20 <- as_tibble(
  data.table::rbindlist(
    lapply(
      acha20,
      subset,
      plan_name == 'TOTAL OF MEDICAID'
    ),
    idcol = 'id'
  )
) |>
  select(
    plan_name, id, contains('21')
  ) |>
  mutate(
    idDate = str_extract(string = id, pattern = '[:digit:]+'),
    year = str_sub(idDate, 1, 4),
    month = str_sub(idDate, 5, 6)
  )
```

Next, I handled those files with differing column lengths. 

The file that had 18 columns omitted certain columns that were in future reports, while the file with 24 columns read in columns that didn't exist.

```{r otherFiles, message = FALSE}
cols_24 <- numColumnsflMedicaid2016to2021 |> filter(number_columns == 24) |>
  select(id) |>
  as_vector()

acha_24 <- readxl::read_excel(path = cols_24, sheet = 'AGEGROUP', skip = 6, col_names = FALSE, col_types = 'text') |>
  select(-c(`...21`, `...22`, `...23`, `...24`)) 

names(acha_24) <- namesAcha20

acha_24_final <- acha_24 |>
  filter(plan_name == 'TOTAL OF MEDICAID') |>
  select(plan_name, contains('21')) |>
  mutate(
    id = cols_24,
    idDate = str_extract(string = cols_24, pattern = '[:digit:]+'),
    year = str_sub(idDate, 1, 4),
    month = str_sub(idDate, 5, 6)
  )

acha_24_final

cols_18 <- numColumnsflMedicaid2016to2021 |> filter(number_columns == 18) |>
  select(id) |>
  as_vector()

acha_18_final <- readxl::read_excel(path = cols_18, sheet = 'AGEGROUP', skip = 6, col_names = namesAcha18, col_types = 'text') |>
  filter(plan_name == 'TOTAL OF MEDICAID') |>
  select(plan_name, contains('21')) |>
  mutate(
    id = cols_18,
    idDate = str_extract(string = cols_18, pattern = '[:digit:]+'),
    year = str_sub(idDate, 1, 4),
    month = str_sub(idDate, 5, 6)
  )

acha_18_final
```

Finally, I combined them all into a single file and created date fields for aggregation/plotting.

```{r singleFile}
flMedicaid2016to2021 <- bind_rows(
  flMedicaidGE20,
  acha_18_final,
  acha_24_final
) |>
  mutate(
    across(.cols = tanf_adult_21_over:duals_21_over, as.numeric),
    yearNum = zoo::as.yearmon(idDate, format = "%Y%m"),
    yearQtr = zoo::as.yearqtr(yearNum, format = "%Y%m")
  ) |> 
  arrange(yearNum)

flMedicaid2016to2021
```

# Start aggregating the ACHA Medicaid Population information

```{r achaAgg}
aggByMonth <- flMedicaid2016to2021 |>
  group_by(yearNum) |>
  mutate(
    total = sum(across(.cols = tanf_adult_21_over:duals_21_over))
  ) |>
  distinct(yearNum, .keep_all = TRUE)

aggByMonth

aggByYearQtr <- flMedicaid2016to2021 |>
  group_by(yearQtr) |>
  mutate(
    total = sum(across(.cols = tanf_adult_21_over:duals_21_over)) / 3
  ) |>
  distinct(yearQtr, .keep_all = TRUE)

aggByYearQtr

write_csv(aggByYearQtr, file = paste0('./data/clean/', format(Sys.Date(), '%Y%m%d'), "_fl-acha-aggregate.csv"), na = '')
```

## Plotting the aggregated ACHA Medicaid Population information

First, aggregate by year and month.

```{r plotAchaAggYearMonth, warning = FALSE, fig.width = 20, fig.height = 12}
plotAggByMonth <- aggByMonth |>
  ggplot() +
  geom_col(aes(x = as.factor(yearNum), y = total, fill = as.factor(yearNum)), color = 'black') +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d() +
  ggtitle("Florida Adult (21+) Medicaid Population by Year and Month") +
  xlab("Month-Year") +
  ylab("Number of Enrollees Per Month") +
  hrbrthemes::theme_ipsum_rc() +
  theme(
    axis.text.x = element_text(size = 14, angle = 90),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14, hjust = 0.5, vjust = -1),
    axis.title.y = element_text(size = 14, hjust = 0.5),
    legend.position = 'none'
  ) +
  labs(
    caption = 'Source: Florida Agency for Healthcare Administration \n
               Download URL: https://ahca.myflorida.com/medicaid/Finance/data_analytics/enrollment_report/'
  )

plotAggByMonth
```

Next, aggregate by year and quarter.

This calculation is done by summing enrollees across three months and then dividing by three.

```{r plotAchaAggYearQtr, warning = FALSE, fig.width = 20, fig.height = 12}
plotAggByYearQtr <- aggByYearQtr |>
  ggplot() +
  geom_col(aes(x = as.factor(yearQtr), y = total, fill = as.factor(yearQtr)), color = 'black') +
  scale_y_continuous(labels = scales::comma) +
  ggtitle("Florida Adult (21+) Medicaid Population by Year and Quarter") +
  xlab("Year-Quarter") +
  ylab("Number of Enrollees per Quarter") +
  scale_fill_viridis_d() +
  hrbrthemes::theme_ipsum_rc() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14, hjust = 0.5, vjust = -1),
    axis.title.y = element_text(size = 14, hjust = 0.5),
    legend.position = 'none'
  ) +
  labs(
    caption = 'Source: Florida Agency for Healthcare Administration \n
               Download URL: https://ahca.myflorida.com/medicaid/Finance/data_analytics/enrollment_report/ \n 
               Quarterly Population Calculation = (Month 1 + Month 2 + Month 3) / 3'
  )

plotAggByYearQtr
```


## Merge SDUD with ACHA Information

```{r sdudAcha}
moudWithFlMedicaid <- yearQuarterAllFLAggregate |>
  inner_join(
    select(aggByYearQtr, yearQtr, total),
    by = c('yearQuarter' = 'yearQtr')
  ) |>
  mutate(ratePer100k = (totalRX / total) * 100000)

moudWithFlMedicaid

write_csv(moudWithFlMedicaid, file = paste0('./data/clean/', format(Sys.Date(), '%Y%m%d'), "_fl-moud-aggregate.csv"), na = '')
```

## Plot the Normalized Rate

```{r moudNormalized100k, fig.width = 20, fig.height = 12}
xPos <- aggByYearQtr |>
  distinct(yearQtr) |>
  rowid_to_column()

xMinimum <- xPos$rowid[xPos$yearQtr == "2020 Q1"]
xMaximum <- xPos$rowid[xPos$yearQtr == "2020 Q2"]

plotMOUDWithFLMedicaid <- moudWithFlMedicaid |>
  ggplot() + 
  geom_point(aes(x = as.factor(yearQuarter), y = ratePer100k, group = allDrugs, color = allDrugs), size = 2) +
  geom_line(aes(x = as.factor(yearQuarter), y = ratePer100k, group = allDrugs, color = allDrugs), size = 1) +
  annotate('rect', xmin = xMinimum, xmax = xMaximum, ymin = -Inf, ymax = Inf, fill = 'red', alpha = 0.1) +
  scale_color_manual(values = c("Bunavail" = "grey", "Buprenorphine" = "#FA841A", "Suboxone" = "#004DA9", "Zubsolv" = "chartreuse4")) +
  xlab('Year - Quarter') +
  ylab('Prescription Rate per 100,000 Adult Medicaid Enrollees') +
  ggtitle("Normalized MOUD Prescription Rate - Florida") +
  theme_ipsum_rc() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 16, hjust = 0.5, vjust = -1),
    axis.title.y = element_text(size = 16, hjust = 0.5),
    legend.position = 'bottom',
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.key.size = unit(3, 'line')
  ) +
  labs(color = 'Drug Name')
  
plotMOUDWithFLMedicaid
```

```{r plotWrite, include = FALSE}
plots <- mget(ls(pattern = 'plot'))

Cairo::CairoPDF(onefile = TRUE, file = paste0('./plots/', format(Sys.Date(), '%Y%m%d'), "_florida-moud-plots.pdf"), width = 18, height = 10)
plots
dev.off()
```